---
title: "Rithinking"
author: "ofer engel"
date: "07/01/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(include = FALSE)
knitr::opts_chunk$set(eval = FALSE)

# Questions and homework etc... 
# Here: https://github.com/rmcelreath/statrethinking_winter2019
# 


library(tidyverse)
library(rstanarm)
library(arm)
library(bayesplot)
library(janitor)
library(metR)
library(rethinking)
library(modelsummary)
precis <- function(x, ...) return(rethinking::precis(x, hist=FALSE, ...))

```

#Chapter 4

```{r cherries, include=FALSE, eval=FALSE}

data("cherry_blossoms")
d <- cherry_blossoms |> na.omit()
precis(d) 


ggplot(d, aes(x=year, y=doy)) + geom_point() + geom_smooth()

knots.n <- 15
knot.list <- quantile(d$year, seq(0,1, length.out=knots.n))

library(splines) 
B <-bs(d$year,
       knots=knot.list[-c(1, knots.n)] ,
       degree=3 ,intercept=TRUE)

qplot(d$year, B[,5])
data.frame(B) |> pivot_longer(cols=c(1:17)) |> 
  cbind(year=rep(d$year, rep(17, length(d$year))))  -> Bs

str(Bs)
data.frame(bf=Bs$name, value=Bs$value, year=Bs$year) |> 
  ggplot() + geom_line(aes(x=year, y=value, color=bf))


m4.7 <- quap(
  alist(
    D  ~ dnorm(mu, sigma), 
    mu ~ a + B %*% w, 
    a ~ dnorm(100,10), 
    w ~ dnorm(0,10), 
    sigma ~ exp(1)
), data = list(D=d$doy, B=B),
start=list( w=rep(0,ncol(B)))) 

rethinking::precis(m4.7, depth=2)


post <-extract.samples(m4.7)
w <-apply(post$w,2,mean)
plot( NULL,xlim=range(d$year),ylim=c(-6,6),
      xlab="year" ,ylab="basis*weight")
for (i in 1:ncol(B)) lines(d$year, w[i]*B[,i], col=i)



mu  <-link(m4.7)$mu
mu_PI <-apply(mu,2, PI, 0.99)
plot( d$year, d$doy, col=col.alpha(rangi2,0.3), pch=16)
shade( mu_PI, d$year, col=col.alpha("black",0.5))
```

## 4E3. Model 
yi ∼ Normal(μ, σ)
μ ∼ Normal(0, 10)
σ ∼ Exponential(1)

$ Pr(\mu, \sigma | Y) = \frac{Pr(Y|\mu, \sigma  ) Pr(\mu)Pr(\sigma)}{\int{Pr(Y|\mu, \sigma)Pr(\mu)Pr(\sigma)\,d\mu\,d\sigma}}$

```{r 4M1}
sigma <- rexp(1e4)
mu <- rnorm(1e4, 0, 10)
y  <- rnorm(1e4, mu, sigma)
qplot(y, geom="histogram")


# 4m2

# m.4m2 <- quap(
#   alist(
#     y  ~ dnorm(mu,sigma),
#     mu ~ dnorm(0.10), 
#     sigma ~ dexp(1)
#     
#   ), data=d
# )




```


## 4M4. Model 
yi ~ Normal(μ, σ)
μ ~ a + b * year
a ~ Normal(0, 50)
b ~ Normal(0,10)
σ ~ Exponential(1)



## 4H1. Model 

```{r 4H1}
data(Howell1)
d <-Howell1
precis(d)

d2 <- d |> filter(age>18) |> 
    mutate(std.weight = (weight-mean(weight))/sd(weight))

qplot(d2$weight, geom="histogram", binwidth=1)
qplot(d2$weight, d2$height)

H.11 <- quap(
  alist(
    height ~ dnorm(mu, sigma), 
    mu  ~ b0 + b1*std.weight, 
    b0   ~ dnorm(178, 20), 
    b1   ~ dlnorm(0, 1),
    sigma ~ dunif(0,50)
  ),   data=d2)

newdata <- data.frame(weight=c(46.95, 43.72, 64.78, 32.59, 54.63)) |> 
  mutate(std.weight = (weight-mean(d2$weight))/sd(d2$weight))

preds <-  link(H.11, newdata |> dplyr::select(std.weight))
data.frame(e.height=apply(preds, 2, mean), 
           PI1=apply(preds,2,PI)[1,], PI2=apply(preds,2,PI)[2,] )

HPDI(preds[,1], prob=.9)

```



## 4H.2

```{r}

d3 <- d |> filter(age<18) |> 
    mutate(std.weight = (weight-mean(weight))/sd(weight))

M4H.2 <- quap(
  alist(
    height <- dnorm(mu, sigma), 
    mu ~ b0+b1*weight, 
    b0 ~ dnorm(178, 25), 
    b1 ~ dnorm(2, 1), 
    sigma ~ dexp(1) 
  ), data=d3)

post <-extract.samples(M4H.2,n=1e4)


qplot(x=weight, y=height, data=d3) + 
  geom_abline(slope=mean(post$b1), intercept=mean(post$b0))

weight.seq <- data.frame(weight=seq(min(d3$weight), max(d3$weight), length.out=30))

sim <- sim(M4H.2, weight.seq)

mu <- link(M4H.2, weight.seq)
qplot(x=weight.seq$weight, y=apply(mu,2,mean), geom="line", color=I("red")) + 
  geom_point(data=d3, aes(weight, height)) + geom_ribbon(aes(x=weight.seq$weight, ymin=apply(mu,2,PI, prob=.99)[1,], ymax=apply(mu,2,PI, prob=.99)[2,]), alpha=.3, fill="red") + 
  geom_ribbon(aes(x=weight.seq$weight, ymin=apply(sim,2,PI, prob=.99)[1,], ymax=apply(sim,2,PI, prob=.99)[2,]), alpha=.4, fill="pink") 

```
For every increase of 10 units in weight, we expect an average increase of 27 units in height. 


## 4H.3



```{r 4H.3}

d4 <- d |>  mutate(std.weight = (weight-mean(weight))/sd(weight), 
           log.weight=log(weight))

M4H.3 <- quap(
  alist(
    height <- dnorm(mu, sigma), 
    mu ~ b0+b1*log.weight, 
    b0 ~ dnorm(178, 25), 
    b1 ~ dlnorm(0, 1), 
    sigma ~ dexp(1) 
  ), data=d4)

precis(M4H.3)
post <-extract.samples(M4H.3,n=1e4)


qplot(x=log.weight, y=height, data=d4) + 
  geom_abline(slope=mean(post$b1), intercept=mean(post$b0))

weight.seq <- data.frame(log.weight=seq(min(d4$log.weight), 
                                    max(d4$log.weight), length.out=30)) |> 
  mutate(weight=round(exp(log.weight), digits = 1))

sim <- sim(M4H.3, weight.seq)

mu <- link(M4H.3, weight.seq)


qplot(x=weight.seq$log.weight, y=apply(mu,2,mean), geom="line", color=I("black"), size=I(.1)) + 
  geom_point(data=d3, aes(log.weight, height)) + geom_ribbon(aes(x=weight.seq$log.weight, ymin=apply(mu,2,PI, prob=.99)[1,], ymax=apply(mu,2,PI, prob=.99)[2,]), alpha=.3, fill="red") + 
  geom_ribbon(aes(x=weight.seq$log.weight, ymin=apply(sim,2,PI, prob=.99)[1,], ymax=apply(sim,2,PI, prob=.99)[2,]), alpha=.4, fill="pink") + 
  scale_x_continuous(name="weight (kg)", breaks=weight.seq$log.weight[seq(1,30,by=3)], 
                     labels = weight.seq$weight[seq(1,30,by=3)])

```

So when we compare someone whose height of say someone 7.4kg (mu=70.84cm) with someone who is 7.4kg*exp(1) = 20.11, (mu=118.81cm) we expect the height to increase by b1=46.91, in fact 118.8-70.84=47.96, which is pretty accurate. 
In other words, adding 1 to the log(x) is like multipying x by e, and the result is that the dependent variable increases by b1=46.91. 

if the dependent variable was log(y), we would say that an increase in one in the independent variable would be like multiplying the y by exp(b1)



```{r}

weight.seq <- data.frame(weight=seq(min(d4$weight), max(d4$weight), length.out=30)) |> mutate(log.weight=log(weight))

preds <- link(M4H.3, weight.seq)
weight.seq$height.hat <- apply(preds, 2, mean)
weight.seq$lower.height <- apply(preds, 2, PI,.99)[1,]
weight.seq$higher.height <- apply(preds, 2, PI,.99)[2,]

sim <- sim(M4H.3, weight.seq)
weight.seq$lower.sim  <- apply(sim, 2, PI,.99)[1,]
weight.seq$higher.sim <- apply(sim, 2, PI,.99)[2,]



ggplot(d4) + 
  geom_point(aes(x=weight, y=height), alpha=.5) + 
  geom_line(data=weight.seq, aes(x=weight,y=height.hat), 
            color="red", size=.1) + 
  geom_ribbon(data=weight.seq, alpha=.5, fill="red", 
              aes(x=weight, 
                  ymin=lower.height, ymax=higher.height))+ 
  geom_ribbon(data=weight.seq, alpha=.4, fill="pink", 
              aes(x=weight, 
                  ymin=lower.sim, ymax=higher.sim))

```



# 4h.5 

Return to data(cherry_blossoms) and model the association between blossom date(doy) and March temperature(temp). Note that there are many missing values in both variables. You may consider a linear model, a polynomial, or aspline on temperature. How well does temperature trend predict the blossom trend?


```{r}
data("cherry_blossoms")
d <- cherry_blossoms
d1 <- d[,2:3] %>% na.omit()
d1$temp_c <- (d1$temp - mean(d1$temp))/sd(d1$temp)

qplot(x=temp_c, y=doy, data=d1) + 
  geom_smooth()

mdl <- quap(
  alist(
    doy ~ dnorm(mu, sigma),
    mu  ~ b0 + b1*temp_c + b2*temp_c^2, 
    b0  ~ dnorm(mean=104,20),
    b1  ~ dnorm(mean=0, sd=5), 
    b2  ~ dnorm(mean=0, sd=1),
    sigma ~ dexp(1)
  ), data=d1)

precis(mdl)

```


# Chapter 5


```{r}
data(milk)
d <- milk

d1 <- d %>% mutate(K=standardize(kcal.per.g), 
                   N=standardize(neocortex.perc), 
                   M=standardize(log(mass)))


qplot(N, K, data=d1)  +
  geom_smooth(formula=y~x)

m5.5.draft <- quap(
  alist(
    K  ~ dnorm(mu, sigma), 
    mu ~ b0 + bN*N,
    b0 ~ dnorm(0,1), 
    bN ~ dnorm(0,1), 
    sigma ~ dexp(1)
    
  ), d1[,9:11] %>% na.omit())

precis(m5.5.draft)

prior.draft <- extract.prior(m5.5.draft)
mu.draft <- link(m5.5.draft, post=prior.draft, data=list(N=c(-2,0,2)))


plot( NULL,xlim=c(-2,2),ylim=c(-2,2))
for (i in 1:50)
  lines(c(-2,0,2),mu.draft[i,],col=col.alpha("black",0.3))


m5.5 <- quap(
  alist(
    K  ~ dnorm(mu, sigma), 
    mu ~ b0 + bN*N,
    b0 ~ dnorm(0,.2), 
    bN ~ dnorm(0,0.5), 
    sigma ~ dexp(1)
    
  ), d1[,9:11] %>% na.omit())

precis(m5.5)


prior <- extract.prior(m5.5)
mu <- link(m5.5, post=prior, data=list(N=c(-2,0,2)))


plot( NULL,xlim=c(-2,2),ylim=c(-2,2))
for (i in 1:50)
  lines(c(-2,0,2),mu[i,],col=col.alpha("black",0.3))


data.frame(b0=prior.draft$b0,
           bN=prior.draft$bN, 
           type="draft") %>% 
  add_row(
           b0=prior$b0,
           bN=prior$bN,
           type="new"
           ) %>%  qplot(x=bN, y=b0, data=., color=type, alpha=I(.3))

data.frame(b0=prior.draft$b0,
           bN=prior.draft$bN, 
           type="draft") %>% 
  add_row(
           b0=prior$b0,
           bN=prior$bN,
           type="new"
           ) %>% ggplot() + 
  geom_histogram(aes(x=bN, fill=type), position="dodge", binwidth=.1)


precis(m5.5)

d1[,9:11] %>% na.omit() -> d2

xseq <- seq(min(d2$N), max(d2$N), length.out=30)
mu <- link(m5.5, data=list(N=xseq))

ggplot() + geom_point(data=d2, aes(x=N, y=K)) + 
  geom_line(aes(x=xseq, y=apply(mu,2,mean))) + 
  geom_ribbon(aes(x=xseq, ymin=apply(mu,2,PI)[1,], ymax=apply(mu,2,PI)[2,]), alpha=.3)

```

Now let's do both predictors together let's see what happens!




```{r}

m5.7 <- quap(
  alist(
    K  ~ dnorm(mu, sigma), 
    mu ~ b0 + bN*N + bM*M,
    b0 ~ dnorm(0,.2), 
    bN ~ dnorm(0,0.5),
    bM ~ dnorm(0,0.5),
    sigma ~ dexp(1)
    
  ), d1[,9:11] %>% na.omit())

precis(m5.7)


m5.7i <- quap(
  alist(
    K  ~ dnorm(mu, sigma), 
    mu ~ b0 + bN*N + bM*M + bMN*M*N,
    b0 ~ dnorm(0,.2), 
    bN ~ dnorm(0,0.5),
    bM ~ dnorm(0,0.5),
    bMN ~ dnorm(0,0.5),
    sigma ~ dexp(1)
    
  ), d1[,9:11] %>% na.omit())

precis(m5.7i)


```

Ok now things with the heights

```{r}
data("Howell1")
d <- Howell1 %>% mutate(male=male+1)

mdl <- quap(
  alist(
    height~dnorm(mu, sigma),
    mu <- a[male], 
    a[male] <- dnorm(178,20),
    sigma ~ dunif(0, 50)      
  ), data=d)

precis(mdl, depth=2)



data("milk")
d <- milk %>% mutate(K = standardize(kcal.per.g), 
                     F = standardize(perc.fat), 
                     L = standardize(perc.lactose))



mdl.KF <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- b0+bF*F, 
    b0 ~ dnorm(0,.3),
    bF ~ dnorm(0,0.5),
    sigma ~ dunif(0, 50)      
  ), data=d)

precis(mdl.KF, depth=2)


mdl.KL <- quap(
  alist(
    K ~ dnorm(mu, sigma),
    mu <- b0+bL*L, 
    b0 ~ dnorm(0,.3),
    bL ~ dnorm(0,0.5),
    sigma ~ dunif(0, 50)      
  ), data=d)

precis(mdl.KF, depth=2)
precis(mdl.KL, depth=2)



```



# Chapter 7

```{r}


set.seed(71) 

# number of plants
N <-100

# simulate initial heights
h0 <-rnorm(N,10,2)

# assign treatments and simulate fungus and growth
treatment <-rep(0:1,each=N/2)
fungus <-rbinom(N,size=1,prob=0.5-treatment*0.4)
h1 <-h0+rnorm(N,5-3*fungus)

# compose a clean data frame
d <-data.frame(h0=h0,h1=h1,treatment=treatment,fungus=fungus)
precis(d)


sim_p <-rlnorm(1e4,0,0.25)

precis( data.frame(sim_p))

mdl6.6 <- quap(
  alist(
    h1 ~ dnorm(mu, sigma), 
    mu <- h0*p, 
    p ~ dlnorm(0, .25), 
    sigma ~ dexp(1)
  ), data=d)

precis(mdl6.1, depth=2)


mdl6.7 <- quap(
  alist(
    h1 ~ dnorm(mu, sigma), 
    mu <- h0*p, 
    p   <-  b0 + bt*treatment + bf*fungus,
    b0  ~   dlnorm(0, .25), 
    bt  ~   dnorm(0, .5),
    bf  ~   dnorm(0, 0.5),
    sigma ~ dexp(1)
  ), data=d)

precis(mdl6.7, depth=2)


mdl6.8 <- quap(
  alist(
    h1 ~ dnorm(mu, sigma), 
    mu <- h0*p, 
    p   <-  b0 + bt*treatment ,
    b0  ~   dlnorm(0, .25), 
    bt  ~   dnorm(0, .5),
    sigma ~ dexp(1)
  ), data=d)

precis(mdl6.8, depth=2)

mdl6.9 <- quap(
  alist(
    h1 ~ dnorm(mu, sigma), 
    mu <- h0*p, 
    p   <-  b0 + bf*fungus ,
    b0  ~   dlnorm(0, .25), 
    bf  ~   dnorm(0, .5),
    sigma ~ dexp(1)
  ), data=d)

precis(mdl6.9, depth=2)


set.seed(77) 


plot(compare( mdl6.6,mdl6.7,mdl6.8, mdl6.9))

plot(compare( mdl6.6, mdl6.8))

```

treatment -> fungus -> h1 <- h0


# Chapter 9 MCMC

```{r}
data(rugged)
d <- rugged
d1 <- d %>% drop_na(rgdppc_2000)  %>% mutate(
  loggdp=log(rgdppc_2000), stdloggdp=loggdp/mean(loggdp),
  std.rugged=rugged/max(rugged), 
  cid = 2-cont_africa) %>% 
  dplyr::select(stdloggdp, std.rugged, cid) 

d2 <- d %>% drop_na(rgdppc_2000)  %>% mutate(
  stdloggdp=log(rgdppc_2000)/mean(log(rgdppc_2000)),
  std.rugged=rugged/max(rugged), 
  cid = 2-cont_africa) %>% 
  dplyr::select(starts_with("stdlo"), std.rugged, cid)

summary(d2)

m8.1 <-quap(
  alist(
    stdloggdp ~dnorm(mu,sigma),
    mu <-a[cid]+b[cid]*(std.rugged-0.215),
    a[cid] ~dnorm(1,0.1),
    b[cid] ~dnorm(0,0.3),
    sigma ~dexp(1)
) ,data=d2)
precis( m8.1,depth=2)



# # summary(dd)
# # 
# dd <- rugged
# dd$log_gdp <-log(d$rgdppc_2000)
# dd <-d[complete.cases(d$rgdppc_2000),]
# dd$log_gdp_std <-dd$log_gdp/mean(dd$log_gdp)
# dd$rugged_std <-dd$rugged/max(dd$rugged)
# dd$cid <-ifelse(dd$cont_africa==1,1,2)
# dd <- dd %>% dplyr::select(log_gdp_std,rugged_std,cid)
# 
# 
# 
# m8.1 <-quap(
#   alist(
#     log_gdp_std ~dnorm(mu,sigma),
#     mu <-a[cid]+b[cid]*(rugged_std-0.215),
#     a[cid] ~dnorm(1,0.1),
#     b[cid] ~dnorm(0,0.3),
#     sigma ~dexp(1)
# ) ,data=dd)
# precis( m8.1,depth=2)
# 
# plot(PSIS(m8.1,pointwise=TRUE)$k)


dat_slim <- list(
  stdloggdp = d2$stdloggdp , 
  stdrugged = d2$std.rugged,
  cid = d2$cid)


m9.1 <- ulam(
  alist(
    stdloggdp ~ dnorm(mu, sigma),
    mu <- b0[cid] + b1[cid]*(stdrugged-0.2149601),
    b0[cid] ~ dnorm(1, 0.2),
    b1[cid] ~ dnorm(0, 0.3), 
    sigma ~ dexp(1)
  ), data=dat_slim, chains=1)


precis( m9.1,depth=2)
show(m9.1)

```









